---
title: "RepData_PeerAssessment1"
author: "robertwcw"
date: "9/7/2020"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
	.Rfliburl <- "https://raw.githubusercontent.com/robertwcw/Rflib/master"
	source(file.path(.Rfliburl,"getRflib.R"),local=TRUE)
	source(getRflib("is.defined.R"),local=TRUE)
	source(getRflib("myplclust.R"),local=TRUE)
	source(getRflib("strCap.R"),local=TRUE)
```

## Introduction

This assignment makes use of the data collected from a personal activity monitoring device strapped to the waist, captures data at 5 minute intervals through out the day on the number of steps taken by an anonymous individual in daily activities during the months of November and October in 2012. The device that was used to capture the steps data is part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. 

The data set in CSV format is available here: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).

Each row in the data set consists of three variables recorded the total number of steps taken in daily activities in every 5 minutes interval spanned across 24 hours each day for two consecutive months. The context of the steps taken in daily activities is referring to the body movements like walking (include up/down stairs), running or jumping.

The variables found in the data set are:  
1. **steps** (Numeric): Number of steps taken by the anonymous individual in 5-minute interval.  
2. **date** (Date): The date (YYYY-MM-DD) which the measurements were taken.  
3. **interval** (Numeric): Identifier for the 5-minute interval for which the observation was captured.

**<data.load>**

1. download data set and unzip to designated directory.  
2. take-a-peek at the unzipped data set to determine number of file, file type, data format and the presence of header...etc, to determine the  appropriate technique to process the 'raw' data into analytic data for downstream analysis.

<!-- as.list(unzip(filetmp, list = TRUE)) #get file list -->
<!-- length(filels$Name) #display number of files in the data set. -->
<!-- filels$Name #display file name & file type. -->
<!-- filels$Length #display physical file size -->
<!-- length(readLines(filels$Name[i])) - 1 #display row count (text-file only) -->
<!-- readLines(filels$Name[i], n = 5) #display first 5 lines. (text-file only) -->

```{r data.load, message = FALSE, results = "hide"}
filetmp <- tempfile()
datadir <- paste(".", "data", sep = "/")
fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileurl, filetmp)

if (!dir.exists(datadir)) 
    {
        dir.create(datadir)
    }

unzip(filetmp, exdir = datadir, overwrite = TRUE)
filels <- as.list(unzip(filetmp, list = TRUE))
filein <- as.character()
filesrc <- as.character()

library(readr)
for (i in 1:length(filels$Name)) 
    {
        filesrc <- paste(datadir, filels$Name[[i]], sep = "/")
        filein <- strsplit(filels$Name[[i]], ".", fixed = TRUE)[[1]][1]
        assign(filein, read_csv(filesrc))
    }
detach("package:readr", unload = TRUE)

unlink(filetmp)
rm(filetmp, fileurl, filein, filesrc, i)
```
  
**<data.eval>**

'activity' data structure.

```{r}
str(activity)
```

The 5-numbers summary of 'activity' data set shows there are 2304 missing values, representing approx. 13.11% missing values out of the entire data set found in the 1st column named **steps** variable, whereas 2nd and 3rd columns **date** and **interval** variables are complete without missing values.

```{r}
summary(activity)
colMeans(is.na(activity))
```


There are some missing columns (on dates Oct01, Oct08, Nov01, Nov04, Nov09, Nov10, Nov14 and Nov30) in the histogram plotted below due to the presence of missing values in the data set.

```{r steps.plot}
names(activity) <- strCap(names(activity))
no.na <- !is.na(activity$Steps)
library(ggplot2)
library(scales)
# ggplot(activity[no.na,]) + geom_col(aes(x = Date, y = Steps, fill = Steps)) +
gr0 <- ggplot(activity[no.na,], aes(x = Date, y = Steps, fill = Steps)) 
gr0 <- gr0 + geom_col() + scale_x_date(breaks = breaks_pretty(11)) 
gr0 <- gr0 + theme(plot.title = element_text(hjust = 0.5),
                   plot.subtitle = element_text(hjust = 0.5),
                   axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
gr0 <- gr0 + labs(title = "Number of Steps Taken for Each Day",
                  subtitle = "(October & November, 2012)")
gr0 <- gr0 + xlab("Date") + ylab("Steps Count")  
print(gr0)
detach("package:ggplot2", unload = TRUE)
detach("package:scales", unload = TRUE)
rm(no.na)
```






**<data.prep>**

```{r data.prep, results = "hide"}
# colnames(activity) <- c("Steps","Date","Interval")
```

k-nearest neighbour imputation method to replace the missing values found in the activity data set.
```{r na.impute}
# library(impute)
# imputed.activity <- impute.knn(data.matrix(activity), k = 15)$data
# detach("package:impute", unload = TRUE)
```

5-numbers summary on 'activity' data set reveals that missing values 'NA' had been imputed with k-nearest neighbour imputation method where k = 15 is used.

```{r}
# summary(activity)
# colMeans(activity)
```







```{r house.keeping}
# detach("package:readr", unload = TRUE)
# detach("package:tidyr", unload = TRUE)
```
